---
title: "Chapter 9 - Multiple and Logistic Regression"
author: "Subhalaxmi Rout"
output:
  html_document:
    df_print: paged
  pdf_document:
    extra_dependencies:
    - geometry
    - multicol
    - multirow
Date: 04/08/2020
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

**Baby weights, Part I.** (9.1, p. 350) The Child Health and Development Studies investigate a range of topics. One study considered all pregnancies between 1960 and 1967 among women in the Kaiser Foundation Health Plan in the San Francisco East Bay area. Here, we study the relationship between smoking and weight of the baby. The variable *smoke* is coded 1 if the mother is a smoker, and 0 if not. The summary table below shows the results of a linear regression model for predicting the average birth weight of babies, measured in ounces, based on the smoking status of the mother.

\begin{center}
\begin{tabular}{rrrrr}
  \hline
            & Estimate  & Std. Error  & t value   & Pr($>$$|$t$|$) \\ 
  \hline
(Intercept) & 123.05    & 0.65        & 189.60    & 0.0000 \\ 
smoke       & -8.94     & 1.03        & -8.65     & 0.0000 \\ 
  \hline
\end{tabular}
\end{center}

The variability within the smokers and non-smokers are about equal and the 
distributions are symmetric. With these conditions satisfied, it is reasonable 
to apply the model. (Note that we don't need to check linearity since the 
predictor has only two levels.)

(a) Write the equation of the regression line.
(b) Interpret the slope in this context, and calculate the predicted birth 
weight of babies born to smoker and non-smoker mothers.
(c) Is there a statistically significant relationship between the average birth weight and smoking?

**Answer**

$(a)$ Regression line euation:
\[\begin{aligned}
\widehat{average\_birth\_weight}= 123.05 + (-8.94) \times smoke
\end{aligned}
\]

$(b)$ Predicted birth weight of babies from 

Smoking mother: 123.05 - 8.94 * 1 = 114.11

Non Smoking mother: 123.05 - 8.94 * 0 = 123.05

$(c)$ Based on hypothesis testing:

H0: No relationship between the average birth weight and smoking 

HA: There is a relationship between average birth and smoking.

From the table, we can see p-value is `0`, which is smaller than 0.05. 
So here reject null hyposthesis against alternative hypothesis. 
We can say there is a relationship between average birth and smoking.


--------------------------------------------------------------------------------

\clearpage

**Absenteeism, Part I.**  (9.4, p. 352) Researchers interested in the relationship between absenteeism from school and certain demographic characteristics of children collected data from 146 randomly sampled students in rural New South Wales, Australia, in a particular school year. Below are three observations from this data set. 

\begin{center}
\begin{tabular}{r c c c c}
  \hline
 	  & eth 	& sex 	& lrn 	& days \\   
  \hline
1 	& 0 		& 1 		& 1 		&   2 \\ 
2 	& 0 		& 1 		& 1 		&  11 \\ 
$\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ & $\vdots$ \\ 
146 & 1 		& 0 		& 0 		&  37 \\ 
  \hline
\end{tabular}
\end{center}

The summary table below shows the results of a linear regression model for predicting the average number of days absent based on ethnic background (`eth`: 0 - aboriginal, 1 - not aboriginal), sex (`sex`: 0 - female, 1 - male), and learner status (`lrn`: 0 - average learner, 1 - slow learner).

\begin{center}
\begin{tabular}{rrrrr}
  \hline
            & Estimate  & Std. Error  & t value   & Pr($>$$|$t$|$) \\ 
  \hline
(Intercept) & 18.93     & 2.57        & 7.37      & 0.0000 \\ 
eth         & -9.11     & 2.60        & -3.51     & 0.0000 \\ 
sex         & 3.10      & 2.64        & 1.18      & 0.2411 \\ 
lrn         & 2.15      & 2.65        & 0.81      & 0.4177 \\ 
  \hline
\end{tabular}
\end{center}

(a) Write the equation of the regression line.
(b) Interpret each one of the slopes in this context.
(c) Calculate the residual for the first observation in the data set: a student who is aboriginal, male, a slow learner, and missed 2 days of school.
(d) The variance of the residuals is 240.57, and the variance of the number of 
absent days for all students in the data set is 264.17. Calculate the $R^2$ and 
the adjusted $R^2$. Note that there are 146 observations in the data set.

**Answer**

$(a)$ Regression line euation:
\[\begin{aligned}
\widehat{days\_absent}= 18.93 + (-9.11) \times eth + (3.10) \times sex + (2.15) \times lrn
\end{aligned}
\]

$(b)$ 

+ The model predicts that non-aboriginal students will have 9.11 less absences.
+ The model predicts that males will be absent 3.1 days more than females
+ The model predicts that slow learners will be absent 2.15 times more than average learners.

$(c)$ Residuals <- 2 - predicted

Lets calculate predicted

```{r}
predicted <- 18.93 + 3.1 + 2.15
residuals <- 2 - predicted
residuals
```

$(d)$ Given 

```{r}
n <- 146 
k <- 3   
residual <- 240.57 
all <- 264.17

r2 <- 1 - (residual/all)  
r_adj <- round(1 - (residual/all) * ((n-1) / (n-k-1)), 6)
paste0("The value of r_adj is ", r_adj)
```


--------------------------------------------------------------------------------

\clearpage

**Absenteeism, Part II.** (9.8, p. 357) Exercise above considers a model that predicts the number of days absent using three predictors: ethnic background (`eth`), gender (`sex`), and learner status (`lrn`). The table below shows the adjusted R-squared for the model as well as adjusted R-squared values for all models we evaluate in the first step of the backwards elimination process. 

\begin{center}
\begin{tabular}{rlr}
  \hline
  & Model               & Adjusted $R^2$ \\ 
  \hline
1 & Full model          & 0.0701 \\ 
2 & No ethnicity        & -0.0033 \\ 
3 & No sex              & 0.0676 \\ 
4 & No learner status   & 0.0723 \\ 
  \hline
\end{tabular}
\end{center}

Which, if any, variable should be removed from the model first?

**Answer**

I would like to remove, `learner status` because this has the highest Adjusted $R^2$ value. 





--------------------------------------------------------------------------------

\clearpage

**Challenger disaster, Part I.** (9.16, p. 380) On January 28, 1986, a routine launch was anticipated for the Challenger space shuttle. Seventy-three seconds into the flight, disaster happened: the shuttle broke apart, killing all seven crew members on board. An investigation into the cause of the disaster focused on a critical seal called an O-ring, and it is believed that damage to these O-rings during a shuttle launch may be related to the ambient temperature during the launch. The table below summarizes observational data on O-rings for 23 shuttle missions, where the mission order is based on the temperature at the time of the launch. *Temp* gives the temperature in Fahrenheit, *Damaged* represents the number of damaged O-rings, and *Undamaged* represents the number of O-rings that were not damaged.

\begin{center}
\begin{tabular}{l rrrrr rrrrr rrrrr rrrrr rrr}
\hline
\vspace{-3.1mm} \\
Shuttle Mission   & 1  & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 \\
\hline
\vspace{-3.1mm} \\
Temperature       & 53 & 57 & 58 & 63 & 66 & 67 & 67 & 67 & 68 & 69 & 70 & 70  \\
Damaged           & 5  & 1 & 1 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
Undamaged         & 1  & 5 & 5 & 5 & 6 & 6 & 6 & 6 & 6 & 6 & 5 & 6 \\
\hline
\\ 
\cline{1-12}
\vspace{-3.1mm} \\
Shuttle Mission   & 13 & 14 & 15 & 16 & 17 & 18 & 19 & 20 & 21 & 22 & 23 \\
\cline{1-12}
\vspace{-3.1mm} \\
Temperature       & 70 & 70 & 72 & 73 & 75 & 75 & 76 & 76 & 78 & 79 & 81 \\
Damaged           & 1  & 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
Undamaged         & 5  & 6 & 6 & 6 & 6 & 5 & 6 & 6 & 6 & 6 & 6 \\
\cline{1-12}
\end{tabular}
\end{center}

(a) Each column of the table above represents a different shuttle mission. Examine these data and describe what you observe with respect to the relationship between temperatures and damaged O-rings.
(b) Failures have been coded as 1 for a damaged O-ring and 0 for an undamaged O-ring, and a logistic regression model was fit to these data. A summary of this model is given below. Describe the key components of this summary table in words.

\begin{center}
\begin{tabular}{rrrrr}
  \hline
            & Estimate & Std. Error & z value   & Pr($>$$|$z$|$) \\ 
  \hline
(Intercept) & 11.6630  & 3.2963     & 3.54      & 0.0004 \\ 
Temperature & -0.2162  & 0.0532     & -4.07     & 0.0000 \\ 
  \hline
\end{tabular}
\end{center}

(c) Write out the logistic model using the point estimates of the model parameters.
(d) Based on the model, do you think concerns regarding O-rings are justified? Explain.


**Answer**

$(a)$ From the table we can see

+ Higher number of damaged O-rings are observed at lower temperature
+ Less number of damaged O-rings are observed at  higher temperature

$(b)$ This model is represented by two components: 

+ Intercept
+ Temperature 

The `estimate` parameter estimate for the model. The z-value and the p-value help with 
distinguishing important information and shows how good the variables predict this model.

$(c)$ Logistic model:

\[\begin{aligned}
log_e(p_i/1 -p_i)= 11.6630 + (-0.2162) \times Temperature
\end{aligned}
\]

$(d)$ yes, based on the model we can say that concerns regarding O-rings are justified. 
There are more damaged O-rings at lower temperatures and O-rings are significant components.




--------------------------------------------------------------------------------

\clearpage

**Challenger disaster, Part II.** (9.18, p. 381) Exercise above introduced us to O-rings that were identified as a plausible explanation for the breakup of the Challenger space shuttle 73 seconds into takeoff in 1986. The investigation found that the ambient temperature at the time of the shuttle launch was closely related to the damage of O-rings, which are a critical component of the shuttle. See this earlier exercise if you would like to browse the original data.

\begin{center}
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.show="hold", out.width="50%", fig.height=4}
library(openintro)
# load data ---------------------------------------------------------
if(!file.exists('orings.rda')) {
	download.file('https://github.com/jbryer/DATA606Fall2019/blob/master/course_data/orings.rda?raw=true',
				  'orings.rda')
}
load("orings.rda")
set.seed(17)
# plot probability of damage vs. temperature ------------------------
these <- orings[,1] %in% c(67, 70, 76)
plot(orings[,1] + 
       c(rep(0, 5), c(-0.1, 0, 0.1), 0, 0, -0.07, -0.07, 0.07, 0.07, 
         rep(0, 4), -0.07, 0.07, 0, 0, 0), 
     orings[,2]/6, 
     xlab = "", ylab = "Probability of damage", 
     xlim = c(50, 82), ylim = c(0,1), 
     col = COL[1,2], pch = 19)
mtext("Temperature (Fahrenheit)", 1, 2)
# probability calculations ------------------------------------------
temperature <- c(51, 53, 55)
logitp <- 11.6630 - 0.2162 * temperature
p <- exp(logitp) / (1+exp(logitp))
```
\end{center}

(a) The data provided in the previous exercise are shown in the plot. The logistic model fit to these data may be written as
\begin{align*}
\log\left( \frac{\hat{p}}{1 - \hat{p}} \right) = 11.6630 - 0.2162\times Temperature
\end{align*}

where $\hat{p}$ is the model-estimated probability that an O-ring will become damaged. Use the model to calculate the probability that an O-ring will become damaged at each of the following ambient temperatures: 51, 53, and 55 degrees Fahrenheit. The model-estimated probabilities for several additional ambient temperatures are provided below, where subscripts indicate the temperature:

\begin{align*}
&\hat{p}_{57} = 0.341
	&& \hat{p}_{59} = 0.251
	&& \hat{p}_{61} = 0.179
	&& \hat{p}_{63} = 0.124 \\
&\hat{p}_{65} = 0.084
	&& \hat{p}_{67} = 0.056
	&& \hat{p}_{69} = 0.037
	&& \hat{p}_{71} = 0.024
\end{align*}

(b) Add the model-estimated probabilities from part~(a) on the plot, then connect these dots using a smooth curve to represent the model-estimated probabilities.
(c) Describe any concerns you may have regarding applying logistic regression in this application, and note any assumptions that are required to accept the model's validity.

**Answer**

$(a)$ 

```{r}
p <- function(temp)
{
  damaged_Oring <- 11.6630 - 0.2162 * temp
  
  pi <- exp(damaged_Oring) / (1 + exp(damaged_Oring))
  
  return (round(pi,5))
}

p51 <- p(51)
paste0("p51: ",p51)
p53 <- p(53)
paste0("p53: ",p53)
p55 <- p(55)
paste0("p55: ",p55)
```

$(b)$

```{r}
temp <- seq(from = 51, to = 81, by = 1)
predicted_prob <- exp(11.6630-(0.2162*temp))/(1+exp(11.6630-(0.2162*temp)))
data <- as.data.frame(cbind(temp, predicted_prob))
graphics::plot(data$temp, data$predicted_prob,col = "steelblue")
```

$(c)$ Key conditions for fitting in Logistic Regression model:

* True relationship between parameter and predictor.
* Outcome for ecah case must be independent of the other outcomes of the other cases.

From the conditions we can assume that these are met. 



